Parameterization of the responses of subarctic European vegetation to key environmental variables for ozone risk assessment

### 2022-03-31 VegOzone.tex
Rephrase Introduction

\begin{abstract}
  The unique vegetation of the subarctic region acclimatized to extremes of cold and midnight sun are likely to be at threat from the combined impacts of climate change and increasing ozone concentrations \chem{[O_3]}. The atmospheric and climatic characteristics of the subarctic are known to lead to pronounced peak \chem{[O_3]} in spring. To date, only a few studies assessed the response of subarctic vegetation to variations in climate and air pollution. With no bespoke parameterization for stomatal response to essential climate variables available, this study looks to fill this knowledge gap by examining an adjustment of existing, Javis type boreal parameterizations, based on local climate condition. To this end, we vary the temperature response function according to the climatological temperature distribution during the subarctic growing season. We evaluate the extent to which two recent years (2018 and 2019) deviate from climatic and \chem{[O_3]} norms and how these potentially more frequent future deviations may influence the projected ozone damage to subarctic vegetation. We find that 2018 was an anomalously warm and bright year, particularly in spring and early summer. Higher than average \chem{[O_3]} was observed in April/May while frequent episodes of ozone volume mixing ratios (VMRs) above $40\,\unit{ppb}$ occurred in June--August. These episodes are in part attributable to forest fires in the Northern Hemisphere and warmer and sunnier conditions. We apply the integrated flux-metric Phytotoxic Ozone Dose (POD) to determine ozone risk and damage to vegetation as a function of \chem{[O_3]}, environmental factors, and species-specific physiology. Our study suggests that using standard boreal parameterizations in assessments likely leads to underestimating the risk of ozone damage in this region. We find that local climate adjusted parameterizations of plant functional types (PFTs) for subarctic vegetation result in an ozone-induced biomass loss of $2.5$ to $17.4\,\unit{\%}$. For some species, this loss is up to $xx\,\unit{\%}$ larger than projected from standard boreal parameterizations used in general assessments. Efforts should be targeted towards accurately defining subarctic vegetations' physiological response to essential climate variables.
\end{abstract}

@@

Ground-level ozone (\chem{O_3}) is a highly toxic pollutant known to cause damage to human health \citep{WHO2008} and to a variety of ecosystems around the World \citep{PT:Emberson2020}. Ozone causes reduced photosynthesis, visible foliar injury, early senescence, and programmed cell death of plants \citep{PCE:Kangasjarvi2005}. Annual global yield losses of four major crops (wheat, rice, maize, and soybean) of $3-15\,\unit{\%}$ have been reported \citep{PJ:Ainsworth2017} and a suggested loss in primary production in forestry of $7\,\unit{\%}$ \citep{GCB:Wittig2009,EP:Matyssek2012} has been attributed to ozone.
The crop yield loss studies indicate a threat to food security in rapidly developing countries, e.g., in East and South-East Asia \citep{GCB:Tang2013,NCC:Tai2014,AE:Chuwah2015,GCB:Mills2018}, where ozone concentrations related to enhanced pollutant emissions from the transport, residential, power generation, and industrial sectors have increased in the past few decades. Even though long-term (pre-1950) observations of ground-level ozone concentrations are scarce; evidence from trend analysis of those few sites suggests that since the industrial revolution, background concentrations in the northern hemisphere have at least doubled and continued to increase \citep{IPCC2013}. Recent trend analyses for Europe indicate that a maximum was reached in 2007 \citep{AE:Derwent2018}, after which tropospheric background ozone started to level off or even decline \citep{ESA:Cooper2014, ACP:Wespes2018,ESA:Gaudel2018}. This can be attributed to a successful implementation of air quality regulations coupled with economic restructuring, reducing the number of episodes of peak concentrations especially in summertime \citep[e.g.,][]{ESA:Fleming2018, ESA:Mills2018}. However, changes in environmental conditions associated with climate change, such as an increase in the frequency of heatwaves, could negate the effectiveness of emission reductions concerning ozone impacts and air quality \citep{NCC:Lin2020}. Tropospheric background ozone concentrations derived from satellite instruments range between $50-65\,\unit{ppb}$ over Fennoscandia but strongly depend on season \citep{ESA:Cooper2014}.

The complex photochemical cycles leading to the formation of \chem{O_3} at ground-level involve precursor gases such as carbon monoxide (\chem{CO}) and hydrocarbons known as volatile organic compounds (VOCs) in the presence of nitrogen oxides (\chem{NO_x}). Despite a relatively short average tropospheric lifetime of approximately $22\,\unit{days}$ \citep{JGR:Stevenson2005,ACP:Young2013}, ozone and its precursors are subject to advection over long distances. Episodes of high ozone concentrations in the absence of local pollution sources can therefore often be attributed to the long-range transport of both ozone and its precursors. In northern Fennoscandia, episodes of elevated ozone concentrations in 2003 and 2006 have been traced back to ozone precursor emissions related to forest fires in southern and eastern Europe, respectively \citep{AE:Lindskog2007,EP:Karlsson2013}. Besides VOCs from anthropogenic sources, hydrocarbons are also emitted by vegetation itself in the form of terpenes and monoterpenes, so-called biogenic volatile substances (BVOCs). These emissions are thought to be a response to factors, such as thermal stress, defense against herbivores, or even attraction of pollinators \citep{TPS:Penuelas2003}.

In the annual cycle in Boreal climates, a distinct maximum of ground-level ozone concentrations typically occurs in spring followed by lower concentrations throughout the summer. Several conditions unique to the Arctic are stimulating the formation of this spring peak. Major pathways of ozone and its precursors' removal from the boundary layer include dry deposition on vegetated surfaces and photochemistry \citep{RG:Clifton2020}. Dry deposition to bare ground or snow and ice-covered surfaces was found to be low \citep{ACP:Helmig2007}. This causes a  substantial suppression of ozone removal \citep{AE:Monks2000} and leads to a build-up of ozone and its precursors during winter. Come spring precursors are photochemically activated again, while snow cover typically continues to prevail and suppress ozone dry deposition.
On the other hand, very high ozone concentrations occurring over shorter periods can often be attributed to deep tropopause folding events over the Arctic during winter. These events facilitate an intrusion of stratospheric air masses which are enriched in ozone \citep{JGR:Skerlak2015}.
The spring peak is more pronounced in northern Fennoscandia than in more southern locations but shows a high interannual variability \citep{AB:Klingberg2009, BER:Klingberg2019}. \citet{ACP:Andersson2017} showed in a modeling study focused on Fennoscandia that the variability of ozone concentrations in winter can be attributed mainly to changes in atmospheric background (transport) of ozone, while summertime abundance is mostly affected by emissions of precursors in the rest of Europe.

@@

A substantial body of evidence exists that suggests flux-based metrics, that relate stomatal ozone uptake to vegetation damage, are biologically more relevant for risk assessments than exposure-based metrics. This is because they can account for particular species characteristics (i.e. physiology and phenology) as well as environmental conditions that can decouple the relationship between ozone concentration, ozone uptake, and consequent damage \citep{PT:Emberson2020}. These flux-based metrics are better suited to represent the actual risk to vegetation from ozone, especially in the climatically extreme parts of Europe \citep{EP:Simpson2007,GCB:Mills2011,ICP:MappingManual2017}. Most of these previous studies have focused on the Mediterranean where soil moisture deficits are thought to be most influential in decoupling concentration from uptake. Relatively few studies have explored the situation in Northern European climates \citep[e.g.][]{iF:Juran2018}.

To estimate the flux-based metric, also referred to as the Phytotoxic Ozone Dose over a threshold y ($\mathrm{POD_y}$), an estimate of the stomatal \chem{O_3} flux ($\Phi_\mathrm{sto}$) is calculated based on the assumption that the concentration of \chem{O_3} at the top of the canopy represents a reasonable estimate of the concentration at the upper surface of the laminar layer for a sunlit upper canopy leaf and environmental factors affecting stomatal conductance. $\mathrm{POD_y}$ is then calculated according to:
%
\begin{equation}
  \mathrm{POD_y} = \int{(\Phi_\mathrm{sto}-y)\cdot \mathrm{dt}},
  \label{eq:pod}
\end{equation}
%
with the hourly averaged stomatal ozone flux $\Phi_\mathrm{sto}$ (see Appendix Eq.~(\ref{eq:flux_stomata}) and a stomatal ozone flux threshold $\mathrm{y}$ both given in units of \unit{nmol\,\chem{O_3}\,m^{-2}\,\mathrm{PLA}\,s^{-1}}. The flux threshold $\mathrm{y}$ represents the detoxification potential of the plant and is typically only exceeded during daylight hours (i.e. when global radiation is above $50\,\unit{W\,m^{-2}}$). $\mathrm{POD_y}$ is given in \unit{mmol\,m^{-2}} integrated over the growing season.

The $\mathrm{POD_y}$ metric is currently used in risk assessments under the United Nations Economic Commission for Europe (UNECE) Long Range Transboundary Air Pollution (LRTAP) Convention to identify those locations across Europe where vegetation is at risk from ozone. This convention aims to develop an effects-based emission reduction policy that can target those ozone precursor emissions that are most influential in causing damage. To achieve this, the concept of critical levels (CLs) has been developed and applied \citep{Maas2016}. Exceedance of the CL is used to identify those areas across Europe that would benefit from targeted Europe-wide emission reductions. Methods to estimate $\mathrm{POD_y}$ and CLs are defined by the UNECE LRTAP Convention and described in a Mapping Manual \citep{ICP:MappingManual2017}. The CL is calculated by: 
%
\begin{equation}
  \mathrm{CL_{exeed}} = \mathrm{POD_y} - \mathrm{CL}.
\end{equation}
%

An application of the Mapping Manual method was made by \citet{GCB:Mills2011} where exceedance of the CL was related to observations of ozone damage to a clover bio-monitor. This study demonstrated the improved ability of flux-based metrics (in comparison to exposure-based metrics) to identify the geographical distribution of the risk of ozone damage and also showed that the ozone damage could extend into more northerly regions of Europe. For our study, we conducted a similar bio-monitoring study at Svanhovd, a site with agrometeorological and air pollution monitoring, to assess whether local ozone concentrations were capable of inducing damage to vegetation growing under northern Fennoscandia conditions. In 2018, we indeed observed ozone damage on clover and tobacco. In contrast, no such damage was found on the clovers in 2019. This indicates that environmental factors in 2018 differed from 2019.

The described methods for risk assessment rely on accurate representation of plant physiological responses to key environmental variables and are most often parameterized using data for less extreme climates, i.e., temperate or continental climates; and to a lesser extent, Boreal and Mediterranean climates. Parameterization of these models for more extreme bio-geographical regions such as subarctic climates has not been performed and relies on assuming that less extreme Boreal parameterization will be applicable. It is important to understand the implications of using these more generic parameterizations for more extreme climate regions. In our study, we explore how risk might be over- or underestimated using generic vs bespoke subarctic parameterization. We also investigate the effects of the interannual variability of essential climate variables (sensu World Meteorological Organization) on future risk of vegetation damage.
This allows an understanding of how the threat from ozone may change in the future and which are the key aspects of plant physiology that might determine the potential risk. Hence, we determine which parameters require special attention for more accurate pollution impact assessment and more generally for describing gas exchange response to a changing climate that will influence biogeochemical cycling. This might also be one key to solve long-standing issues of earth system models (ESM) which often lack PFTs specifically representing subarctic conditions in their land surface models \citep{GMD:Poulter2015,JAMES:Lawrence2019}.

The observation site Svanhovd and available observational data are presented in Section~\ref{sec:data}. From these data, we derive multi-annual averages of essential climate variables (temperature, precipitation, global irradiance, and ozone) and evaluate the deviation of 2018/19 from this norm (Section~\ref{sec:stats}). In Section~\ref{sec:do3se}, we use the $\mathrm{DO_3SE}$ model to estimate the ozone uptake by natural and semi-natural vegetation to investigate the main environmental drivers for projected ozone damage risk in 2018. To this end, we derive bespoke parameterizations for subarctic species to assess over- and underestimations arising from the use of generic parameterizations established from less climatically extreme conditions. We will summarize our results in Section~\ref{sec:conc} and discuss how this research can support improvements in future assessments of pollution risk and more generally biogeochemical cycling.

@@

>>>>

\begin{abstract}
  The unique vegetation of the subarctic region acclimatized to extremes of cold and midnight sun are likely to be at threat from the combined impacts of climate change and increasing ozone concentrations \chem{[O_3]}. The atmospheric and climatic characteristics of the subarctic are known to lead to pronounced peak \chem{[O_3]} in spring. To date, only a few studies assessed the response of subarctic vegetation to variations in climate and air pollution. This study looks to fill this knowledge gap by examining the uncertainty in risk assessments which are usinge more general boreal paramterizations. With no bespoke parameterization for stomatal response to essential climate variables available, we propose an adjustment of existing, Javis type boreal parameterizations, based on local climate condition. To this end, we vary the temperature response function according to the climatological temperature distribution during the subarctic growing season. We evaluate the extent to which two recent years (2018 and 2019) deviate from climatic and \chem{[O_3]} norms and how these potentially more frequent future deviations may influence the projected ozone damage to subarctic vegetation. We find that 2018 was an anomalously warm and bright year, particularly in spring and early summer. Higher than average \chem{[O_3]} was observed in April/May while frequent episodes of ozone volume mixing ratios (VMRs) above $40\,\unit{ppb}$ occurred in June--August. These episodes are in part attributable to forest fires in the Northern Hemisphere and warmer and sunnier conditions. We apply the integrated flux-metric Phytotoxic Ozone Dose (POD) to determine ozone risk and damage to vegetation as a function of \chem{[O_3]}, environmental factors, and species-specific physiology. Our study suggests that using standard boreal parameterizations in assessments likely leads to underestimating the risk of ozone damage in this region. We find that local climate adjusted parameterizations of plant functional types (PFTs) for subarctic vegetation result in an ozone-induced biomass loss of $2.5$ to $17.4\,\unit{\%}$. For some species, this loss is up to $xx\,\unit{\%}$ larger than projected from standard boreal parameterizations used in general assessments. Efforts should be targeted towards accurately defining subarctic vegetations' physiological response to essential climate variables.
\end{abstract}

\newpage

@@

Ground-level ozone (\chem{O_3}) is a highly toxic pollutant known to cause damage to a variety of ecosystems around the World \citep{PT:Emberson2020}. Ozone causes reduced photosynthesis, visible foliar injury, early senescence, and programmed cell death of plants \citep{PCE:Kangasjarvi2005}. Hence affecting the global carbon cycle \citep{BGS:Lombardozzi2013} and crop yields. Annual global yield losses of four major crops (wheat, rice, maize, and soybean) of $3-15\,\unit{\%}$ have been reported \citep{PJ:Ainsworth2017} and a suggested loss in primary production in forestry of $7\,\unit{\%}$ \citep{GCB:Wittig2009,EP:Matyssek2012} has been attributed to ozone. Even though long-term (pre-1950) observations of ground-level ozone concentrations are scarce; evidence from trend analysis of those few sites suggests that since the industrial revolution, background concentrations in the northern hemisphere have at least doubled and continued to increase \citep{IPCC2013}. Recent trend analyses for Europe indicate that a maximum was reached in 2007 \citep{AE:Derwent2018}, after which tropospheric background ozone started to level off or even decline \citep{ESA:Cooper2014, ACP:Wespes2018,ESA:Gaudel2018}. This can be attributed to a successful implementation of air quality regulations coupled with economic restructuring, reducing the number of episodes of peak concentrations especially in summertime \citep[e.g.,][]{ESA:Fleming2018, ESA:Mills2018}. However, changes in environmental conditions associated with climate change, such as an increase in the frequency of heatwaves, could negate the effectiveness of emission reductions concerning ozone impacts and air quality \citep{NCC:Lin2020}. Tropospheric background ozone concentrations derived from satellite instruments range between $50-65\,\unit{ppb}$ over Fennoscandia but strongly depend on season \citep{ESA:Cooper2014}.

In the annual cycle, a distinct maximum of ground-level ozone concentrations typically occurs in spring followed by lower concentrations throughout the summer. This ozone spring peak is more pronounced in northern Fennoscandia than in more southern locations but shows a high interannual variability \citep{AB:Klingberg2009, BER:Klingberg2019}. \citet{ACP:Andersson2017} showed in a modeling study focused on Fennoscandia that the variability of ozone concentrations in winter can be mainly attributed to changes in atmospheric background (transport) of ozone, while summertime abundance is mostly affected by emissions of precursors in the rest of Europe.

The precursor gases involved in the formation of ground-level ozone are carbon monoxide (\chem{CO}) and hydrocarbons known as volatile organic compounds (VOCs) in the presence of nitrogen oxides (\chem{NO_x}). In northern Fennoscandia, episodes of elevated summertime ozone concentrations in 2003 and 2006 have been traced back to ozone precursor emissions related to forest fires in southern and eastern Europe, respectively \citep{AE:Lindskog2007,EP:Karlsson2013}. photosynthetically active vegetation in the form of terpenes and monoterpenes, so-called biogenic volatile substances (BVOCs) and soil. These emissions are thought to be a response to factors, such as thermal stress, defense against herbivores \citep{Ecosystems:Ghimire2021}, or even attraction of pollinators \citep{TPS:Penuelas2003}. \citet{Ecosystems:Ghimire2021} found no clear evidence that general warming would increase the BVOC emissions in subarctic vegetation, while a reduction of herbivory would decrease them by $44-72\,\%$.

@@

Flux-based metrics, that relate stomatal ozone uptake to vegetation damage, are biologically more relevant for risk assessments than exposure-based metrics because they can account for particular species characteristics (i.e. physiology and phenology) as well as environmental conditions that can decouple the relationship between ozone concentration, ozone uptake, and consequent damage \citep{PT:Emberson2020}. These flux-based metrics are better suited to represent the actual risk to vegetation from ozone, especially in the climatically extreme parts of Europe \citep{EP:Simpson2007,GCB:Mills2011,ICP:MappingManual2017}. But no bespoke parameterization for subarctic vegetation is available.

To estimate the flux-based metric, also referred to as the Phytotoxic Ozone Dose over a threshold y ($\mathrm{POD_y}$), an estimate of the stomatal \chem{O_3} flux ($\Phi_\mathrm{sto}$) is calculated based on the assumption that the concentration of \chem{O_3} at the top of the canopy represents a reasonable estimate of the concentration at the upper surface of the laminar layer for a sunlit upper canopy leaf and environmental factors affecting stomatal conductance. $\mathrm{POD_y}$ is then calculated according to:
%
\begin{equation}
  \mathrm{POD_y} = \int{(\Phi_\mathrm{sto}-y)\cdot \mathrm{dt}},
  \label{eq:pod}
\end{equation}
%
with the hourly averaged stomatal ozone flux $\Phi_\mathrm{sto}$ (see Appendix Eq.~(\ref{eq:flux_stomata}) and a stomatal ozone flux threshold $\mathrm{y}$ both given in units of \unit{nmol\,\chem{O_3}\,m^{-2}\,\mathrm{Projected Leaf Area (PLA)}\,s^{-1}}. The flux threshold $\mathrm{y}$ represents the detoxification potential of the plant and is typically only exceeded during daylight hours (i.e. when global radiation is above $50\,\unit{W\,m^{-2}}$). $\mathrm{POD_y}$ is given in \unit{mmol\,m^{-2}} integrated over the growing season.

The $\mathrm{POD_y}$ metric is currently used in risk assessments under the United Nations Economic Commission for Europe (UNECE) Long Range Transboundary Air Pollution (LRTAP) Convention to identify those locations across Europe where vegetation is at risk from ozone. This convention aims to develop an effects-based emission reduction policy that can target those ozone precursor emissions that are most influential in causing damage. To achieve this, the concept of critical levels (CLs) has been adopted and applied \citep{Maas2016}. Exceedance of the CL is used to identify those areas across Europe that would benefit from targeted Europe-wide emission reductions. Methods to estimate $\mathrm{POD_y}$ and CLs are defined by the UNECE LRTAP Convention and described in a Mapping Manual \citep[most recent version]{ICP:MappingManual2017}. The exceedance of the CL is calculated by: 
%
\begin{equation}
  \mathrm{CL_{exeed}} = \mathrm{POD_y} - \mathrm{CL}.
\end{equation}
%

An application of the Mapping Manual method was made by \citet{GCB:Mills2011} where exceedance of the CL was related to observations of ozone damage to a clover bio-monitor. This study demonstrated the improved ability of flux-based metrics (in comparison to exposure-based metrics) to identify the geographical distribution of the risk of ozone damage and also showed that the ozone damage could extend into more northerly regions of Europe.

The described methods for risk assessment rely on accurate representation of plant physiological responses to key environmental variables and are most often parameterized using data for less extreme climates, i.e., temperate or continental climates; and to a lesser extent, Boreal and Mediterranean climates. Parameterization of these models for more extreme bio-geographical regions such as subarctic climates has not been performed and relies on assuming that less extreme Boreal parameterization will be applicable. It is important to understand the implications of using these more generic parameterizations for more extreme climate regions. In our study, we explore how risk might be over- or underestimated using Boreal parameterization and explore an adjustment of Javis type temperature response functions based on climate data. This allows an understanding of how the threat from ozone may change in the future and which are the key aspects of plant physiology that might determine the potential risk. Our results may hint a solution to solve long-standing issues of earth system models (ESM) which often lack PFTs specifically representing subarctic conditions in their land surface models and display over- or underproductivity in the Arctic \citep{GMD:Poulter2015,JAMES:Lawrence2019}.

Data and methods including the DO3SE model are described in Section 2. The model experiments and derived parameterizations are summarized in Section 3. In Section 4, we present our modelling results. We discuss our results and give an outlook on how these could improve future assessments of pollution risk.

@@

For our study, we conducted a similar bio-monitoring study at Svanhovd, a site with agrometeorological and air pollution monitoring, to assess whether local ozone concentrations were capable of inducing damage to vegetation growing under northern Fennoscandia conditions. In 2018, we indeed observed ozone damage on clover and tobacco. In contrast, no such damage was found on the clovers in 2019. This indicates that environmental factors in 2018 differed from 2019.
